---
title: "SSDB"
author: "William Brown"
date: "1/19/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 10, fig.height = 5)
```

# Loading libraries
```{r echo=TRUE}
library(readxl)
library(tidyverse)
library(lubridate)
library(ggtext)
library(extrafont)
```

# Reading in the data
Data comes
```{r echo=TRUE}
incident_raw <- read_xlsx('SSDB_Raw_Data.xlsx', sheet = 2)
shooter_raw <- read_xlsx('SSDB_Raw_Data.xlsx', sheet = 3)
victim_raw <- read_xlsx('SSDB_Raw_Data.xlsx', sheet = 4)
weapon_raw <- read_xlsx('SSDB_Raw_Data.xlsx', sheet = 5)
```

# Basic Data transformation
The data came in four disparate datasets partaining to shootings, shooter profiles, victim data, and weapons data. While more robust than other data sources, there are serious data quality issues present. For example, the most common age provided in the `shooter` dataset is N/A (i.e., 16% of entries) and 89% of all racial characteristics provided in the `victims` dataset were N/A. Accordingly, data has been filtered and sorted without undergoing imputation. The primary transformation tasks included pivoting, mutating and aggregating data, as well as renaming columns for standardization.

```{r echo = TRUE}
incident <- incident_raw %>%
  rename_all(., tolower) %>%
  mutate(
    year = year(as_date(date)),
    month = month(as_date(date)),
    day = day(as_date(date)),
    day_of_week = weekdays(as_date(date)),
    first_shot = format(strptime(first_shot, "%I:%M %p"), format="%H:%M:%S"),
    first_shot = if_else(first_shot == '', 'N/A', first_shot),
    ) %>%
  mutate(incident_time = if_else(is.na(first_shot), 'N/A', first_shot)) %>%
  select(-c(sources, number_news, date, quarter, media_attention, reliability))

shooter <- shooter_raw %>%
  select(-c(criminalhistory, minorchargedadult, verdict)) %>%
  mutate(age = as.numeric(age)) %>%
  mutate(age_group = case_when(age < 13 ~ 'Child',
                               13 <= age & age <= 17 ~ 'Teen',
                               age >= 18 ~ 'Adult')) %>%
  rename(incident_id = 1) %>%
  count(incident_id, shooterdied) %>%
  pivot_wider(names_from = shooterdied, values_from = n) %>%
  rename(shooter_survived = 2,
         shooter_died = 3,
         shooter_na = 4) %>%
  mutate_all(~replace_na(., 0))

weapon <- weapon_raw %>%
  select(-weapondetails) %>%
  mutate(weapontype = tolower(str_replace_all(weapontype, ' ', '_'))) %>%
  rename(incident_id = 1) %>%
  count(incident_id, weapontype) %>%
  pivot_wider(names_from = weapontype, values_from = n) %>%
  mutate_all(., ~replace_na(., 0)) %>%
  mutate(handguns = handgun + mulitiple_handguns + multiple_handguns,
         rifles = rifle + multiple_rifles,
         other = no_data + multiple_unknown + unknown,
         total_weapons = handguns + rifles + other) %>%
  select(-c(handgun, mulitiple_handguns, multiple_handguns, rifle, multiple_rifles, no_data, multiple_unknown, unknown))

victim_count <- victim_raw %>%
  rename(incident_id = 1) %>%
  count(incident_id, injury) %>%
  pivot_wider(names_from = injury, values_from = n) %>%
  mutate_at(vars('Fatal', 'Wounded', 'None', 'Minor Injuries', 'NA'), ~replace_na(.,0)) %>%
  rename_all(., tolower)

ss_df <- left_join(x = incident, y = shooter, by = 'incident_id') %>%
  left_join(., victim_count, by = 'incident_id') %>%
  left_join(., weapon, by = 'incident_id')
```

# Basic EDA and Visualizaiton
The first visualization looked atthe total number of shootings over time. As it seems, there is a constant upward trend since 1970, the first date in the group. It is important to note that 2021 shootings were not included due to the COVID-19 pandemic and its potential impact on shooting frequency. A second graph looks at the total number of shootings by type, and bins them into five categories (including "Other").

```{r echo = TRUE}
ss_df %>%
  count(year) %>%
  filter(year != '2021') %>%
  ggplot(aes(x = year, y = n)) + 
  annotate(geom = 'text', x = 2002, y = 120, label = 'Max number of shootings\nsince data was collected', color = 'Black', family = 'Times', size = 4, hjust = 0) + 
  annotate(geom = 'segment', x = 2010, xend = 2017, y = 119, yend = 119, color = 'Black') +
  geom_line(size = 1, color = '#3A4869', alpha = 0.8, width = .2) + 
  geom_point(color = '#3A4869', size = 3, shape = 16) + 
  labs(x = 'YEAR',
       y = 'NUMBER OF SCHOOL SHOOTINGS',
       title = 'SCHOOL SHOOTINGS',
       subtitle = 'Since 1970, there have been 1,641 school shootings in the United States.') + 
  theme(panel.background = element_blank(),
        panel.grid = element_line(linetype = 3, color = '#dbdbdb'),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(family = 'Andale Mono', size = '12', color = '#423E37'),
        axis.title.x = element_text(family = 'Andale Mono', size = '12', colour = '#423E37'),
        axis.title.y = element_text(family = 'Andale Mono', size = '12', colour = '#423E37'),
        plot.title = element_text(family = 'Andale Mono', size = '32', colour = 'Black'),
        plot.subtitle = element_text(family = 'Times', size = '14', colour = 'Black', face = 'italic'))

ss_df %>%
  filter(!is.na(situation)) %>%
  mutate(situation = fct_lump(situation, n = 4)) %>%
  count(year, situation, preplanned) %>%
  filter(year != '2021') %>%
  ggplot(aes(x = year, y = n)) + 
  facet_wrap(~situation, nrow = 1) +
  geom_col(fill = '#3A4869', size = 1, color = '#3A4869') +
  labs(x = 'YEAR',
       y = 'NUMBER OF SCHOOL SHOOTINGS',
       title = 'SCHOOL SHOOTINGS BY TYPE') +
  theme(panel.background = element_blank(),
        panel.grid = element_line(linetype = 3, color = '#dbdbdb'),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(family = 'Andale Mono', size = '12', color = '#423E37'),
        axis.text.x = element_text(angle = 270),
        axis.title.x = element_text(family = 'Andale Mono', size = '12', colour = '#423E37'),
        axis.title.y = element_text(family = 'Andale Mono', size = '12', colour = '#423E37'),
        plot.title = element_text(family = 'Andale Mono', size = '20', colour = 'Black'),
        plot.subtitle = element_text(family = 'Times', size = '14', colour = 'Black', face = 'italic'),
        strip.background = element_blank(),
        strip.text = element_text(family = 'Andale Mono'),
        panel.spacing = unit(.5, 'cm'))
```

